---
title: "Learning Imputation"
author: "Eric R. Scott"
date: "2020-07-21"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(tsibble)
library(imputeTestbench)
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Explore options for imputation of missing values

# Load & Wrangle Data

```{r data, echo=TRUE}
bdffp <- read_csv(here("data_cleaned", "daily_precip.csv"))
```

I think I want this wide with just date, and site columns with precip values.  I'll remove ALL accumulations

```{r}
bdffp_wide <-
  bdffp %>% 
  mutate(precip = ifelse(is.na(lag(precip)), NA, precip)) %>% 
  as_tsibble(key = site, index = date) %>% 
  fill_gaps() %>% 
  select(date, site, precip) %>% 
  as_tibble() %>% 
  pivot_wider(names_from = site, values_from = precip)
```

```{r}
manaus <- read_csv(here("data_cleaned", "manaus_weather.csv"))
manaus <-
  manaus %>%
  filter(between(date, min(bdffp$date), max(bdffp$date))) %>% 
  select(date, precip) %>%
  as_tsibble(index = date) %>% 
  as.ts()
```

# Generate test data

```{r}
manaus_test <- sample_dat(manaus, smps = "mar", blck = 7, blckper = FALSE)
```

# Linear Regression

Use 4 bdffp sites with most complete observations.  Join to test data

```{r}
test_ts <- manaus_test[[1]]

df <- left_join(bdffp_wide, as_tsibble(test_ts), by = c("date" = "index")) %>% 
  select(date, value, dimona, `porto alegre`, colosso, km41)
```


fit a regression with only complete cases.

```{r}
df_complete <- df %>% filter(complete.cases(.))
m <- lm(value ~ dimona  + colosso + km41, data = df_complete, na.action = "na.exclude")
coef(m)
```
Oh, wait, duh.  That is the only way it works anyways.  Fuck.

Now, use predict to fill in NAs?

```{r}
pred <- broom::augment(m, data = df, newdata = df)
pred %>% filter(is.na(.fitted))
pred %>% filter(is.na(value) & !is.na(.fitted))
```
Only imputes 95 values because only 95 rows where there is data for the three other sites.


# Multivariate imputation

What about the `mice` package?  Can I treat preip data at each site (or even site:year?) like multivariate data and impute from that?

```{r}
library(mice)
```
```{r}
df <- left_join(bdffp_wide, as_tsibble(test_ts), by = c("date" = "index")) %>% janitor::clean_names()
x1 <- mice(df, maxit = 20)


plot(x1)
stripplot(x1, dimona ~.imp)
x1$method
summary(x1)
```

```{r}
complete(x1, "long") %>%
  group_by(.id) %>%
  summarize_all(mean) %>%
  pull(value)
```


# Check against true data

Use functions in imputeTestbench to calculate RMSE and stuff
