---
title: "Comparison of precipitation and SPEI from different data sources"
author: "Eric R. Scott"
date: "Date Initiated"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(tsibble)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
source(here("R", "plotting.R"))
```

*Last compiled: `r Sys.Date()`*

# Purpose

1. Compare rainfall from daily gridded datasets with (non-missing, non-accumulated) values from rain gauges at BDFFP.
2. Compare monthly SPEI values calculated with imputed rain gauge data, gridded datasets, and from TRMM and GPM.

# Comparison of daily rainfall data

```{r data, echo=TRUE}
bdffp <- read_csv(here("data_cleaned", "daily_precip.csv"))
imp <- read_csv(here("data_complete", "daily_imputed.csv"))
sa <- read_csv(here("data_cleaned", "sa_daily_1x1.csv"))
xa <- read_csv(here("data_cleaned", "xavier_daily_0.25x0.25.csv"))
manaus <- read_csv(here("data_raw", "embrapa", "Estacao_Manaus_1980-01-01_2016-12-31.csv"))
rpde <- read_csv(here("data_raw", "embrapa", "Estacao_Rio Preto da Eva_1980-01-01_2016-12-31.csv"))
```

Raw BDFFP rain gauge data, with accumulations removed

```{r}
bdffp_wide <-
  bdffp %>% 
  filter(flag != "A") %>% 
  select(site, date, precip) %>% 
  pivot_wider(names_from = site, values_from = precip) %>% 
  arrange(date) %>% 
  rowwise() %>% 
  mutate(bdffp = mean(c_across(-date), na.rm = TRUE)) %>% 
  ungroup()

bdffp_ts <-
  bdffp_wide %>% 
    as_tsibble(index = date) %>% 
    fill_gaps()
```
Gridded data from south america

```{r}
sa2 <-
  sa %>% 
  #only one grid cell
  select(date, sa = precip)

daily_sources <- left_join(bdffp_ts, sa2, by = "date")
```

Gridded data from Xavier et al.

```{r}
xa_wide <- 
  xa %>% 
  unite(latlon, lat, lon, sep = ", ") %>% 
  select(date, latlon, precip) %>% 
  pivot_wider(names_from = latlon, names_prefix = "xa_", values_from = precip) %>% 
  rowwise() %>% 
  mutate(xa = mean(c_across(-date), na.rm = TRUE)) %>% 
  ungroup()

daily_sources2 <- left_join(daily_sources, xa_wide, by = "date")
```

Weather stations

```{r}
stations <- 
  full_join(
  manaus %>% 
  select(date = Data,
         manaus = Precipitacao),
  rpde %>% 
    select(date = Data,
           rpde = Precipitacao),
  by = "date")
daily_sources3 <- left_join(daily_sources2, stations, by = "date")
```

Imputed daily data

```{r}
imp_wide <- imp %>% 
  pivot_wider(names_from = site, names_prefix = "imp_", values_from = precip)
daily_sources4 <- left_join(daily_sources3, imp_wide, by = "date")
```


# Calculate correlations

Using all pairwise complete observations and non-parameteric Spearman correlation.

```{r}
cor_mat <- 
  daily_sources4 %>% 
  as_tibble() %>% 
  select(-date) %>% 
  cor(use = "pairwise.complete.obs", method = "spearman") 

cor_df <-
  cor_mat %>%
  # cor() %>% 
  as_tibble(rownames = "source_a") %>% 
  pivot_longer(-source_a, names_to = "source_b", values_to = "cor") %>% 
  mutate(source_a = fct_inorder(source_a),
         source_b = fct_inorder(source_b))

ggplot(cor_df, aes(x = source_a, y = source_b)) +
  geom_point(aes(size = abs(cor), color = cor)) +
  scale_color_continuous(type = "viridis") +
  theme_bw() +
  theme(axis.text.x.bottom = element_text(angle = 45, hjust =1),
        axis.title = element_blank()) +
  labs(title = "Pairwise correlations among daily precipitation data sources",
       caption = "bdffp = mean of rain gauges\n sa and xa are gridded data products\n manaus and rpde are weather stations\n imp_* are imputed values using all these data")
```
Why are some of the rain gauges so strongly negatively correlated?

```{r}
ggplot(daily_sources4, aes(x = florestal, y = km41)) +
  geom_point()
ggplot(daily_sources4, aes(x = florestal, y = `porto alegre`)) +
  geom_point()
ggplot(daily_sources4, aes(x = gaviao, y = `cabo frio`)) +
  geom_point()
```

Because there's only like 3 days when there is data from both!

# Monthly rainfall data

For this, I guess I'll keep all accumulations

```{r}
bdffp_all <-
  bdffp %>% 
  # filter(flag != "A") %>% 
  select(site, date, precip) %>% 
  pivot_wider(names_from = site, values_from = precip) %>% 
  arrange(date) %>% 
  rowwise() %>% 
  mutate(bdffp = mean(c_across(-date), na.rm = TRUE)) %>% 
  ungroup() %>% 
  select(date, bdffp)

bdffp_ts2 <-
  bdffp_all %>% 
    as_tsibble(index = date) %>% 
    fill_gaps()

monthly_sources <- 
  full_join(bdffp_ts2, 
          daily_sources4 %>% 
            select(-c(km37:bdffp))) %>% 
  mutate(yearmonth = yearmonth(date), .after = date) %>% 
  as_tibble() %>%
  group_by(yearmonth) %>% 
  summarise(across(-c(date), ~sum(., na.rm = TRUE), .names = "{col}")) %>% 
  ungroup()
```

```{r}
cor_mat2 <- 
  monthly_sources %>% 
  as_tibble() %>% 
  select(-yearmonth) %>% 
  cor(use = "pairwise.complete.obs", method = "spearman") 

cor_df2 <-
  cor_mat2 %>%
  # cor() %>% 
  as_tibble(rownames = "source_a") %>% 
  pivot_longer(-source_a, names_to = "source_b", values_to = "cor") %>% 
  mutate(source_a = fct_inorder(source_a),
         source_b = fct_inorder(source_b)) %>% 
  mutate(cor = ifelse(source_a == source_b, NA, cor)) #remove diagonal

ggplot(cor_df2, aes(x = source_a, y = source_b)) +
  geom_point(aes(size = abs(cor), color = cor)) +
  scale_color_continuous(type = "viridis") +
  theme_bw() +
  theme(axis.text.x.bottom = element_text(angle = 45, hjust =1),
        axis.title = element_blank()) +
  labs(title = "Pairwise correlations for monthly aggregated precipitation",
       caption = "'bdffp' = mean of rain gauges\n 'sa' and 'xa' are gridded data products\n 'manaus' and 'rpde' are weather stations\n 'imp_*' are imputed using all these data")
```
# SPEI data

Here I'll compare SPEI data calculated using the imputed data, the `xa` dataset, and values retrieved from TRMM and GPM.

Eventually/ideally, I'll do this for every plot, but to start with, I'll just use the colosso cluster of plots.

## Read in imputed SPEI

```{r}
spei_imp <- read_csv(here("data_complete", "mon_precip_spi_imputed.csv"))
spei_imp <- spei_imp %>% filter(site == "colosso_clust") %>% select(date, imp_spei = spei) 
spei_imp
```
## Calculate SPEI with `xa` dataset

Which grid cell is Colosso in?  Center is at -59.6, -2.4, roughly.

```{r}
xa_mon <-
  xa %>%
  unite(latlon, lat, lon, sep = ", ") %>% 
  filter(latlon == "-2.375, -59.625") %>% 
  select(date, precip, eto) %>% 
  mutate(yearmonth = yearmonth(date)) %>% 
  group_by(yearmonth) %>% 
  summarize(across(c(precip, eto), sum)) %>% 
  as_tsibble(index = yearmonth)
```
```{r}
library(SPEI)
spei_xa <-
  xa_mon %>% 
  mutate(cb = precip - eto) %>% #calculate climatic water balance
  mutate(xa_spei = as.numeric(spei(.$cb, scale = 3)$fitted)) %>% 
  mutate(date = as.Date(yearmonth)) %>% 
  as_tibble() %>% 
  select(date, xa_spei)
spei_xa
```
# Get pre-calculated SPEI data

From Global SPEI database (spei.csic.es/database.html)

```{r}
global <- read_delim(here("data_raw", "global_spei_-59.75_ -2.25.csv"), delim = ";")
spei_global <-
  global %>% 
  rename(date = "days since 1900-1-1", g_spei = spei)

range(spei_global$date)
```

# Join and compare

actually easier to join by yearmonth

```{r}
spei_global <- spei_global %>% mutate(yearmonth = yearmonth(date), date = NULL)
spei_xa <- spei_xa %>% mutate(yearmonth = yearmonth(date), date = NULL)
spei_imp <- spei_imp %>% mutate(yearmonth = yearmonth(date), date = NULL)

spei_all <-
  inner_join(inner_join(spei_imp, spei_global), spei_xa) %>% 
  select(yearmonth, everything())

spei_all %>% 
  select(-yearmonth) %>% 
  cor(use = "pairwise.complete.obs")
```

Huh, wow.  Not great.

```{r}
df <- spei_all %>% 
  filter(complete.cases(.)) %>% 
  pivot_longer(-yearmonth, names_to = "source", values_to = "spei_3") %>% 
  filter(lubridate::year(yearmonth) != 1987)

p <- ggplot(df, aes(x = yearmonth, y = spei_3, color = source)) +
  geom_line() +
  scale_x_yearmonth("Date", expand = c(0,0)) +
  facet_wrap(~lubridate::year(yearmonth), scales = "free_x")
annotate_spei(p)
```

So direction of drought mostly the same, magnitude fairly different from global SPEI database sometimes. (2004 and 2009).


