---
title: "Precipitation Data EDA"
author: "Eric R. Scott"
date: "2020-07-02"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(here)
library(conflicted)
library(plotly)
library(lubridate)
library(tsibble)
library(corrr)
conflict_prefer("filter", "dplyr") 
conflict_prefer("lag", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose
Identify problems in the data.  Figure out what to do with missing values and multi-day accumulated values. Explore consistency of precip data across sites at BDFFP at different levels of aggregation (daily, monthly, yearly). 

# Load Data

```{r data, echo=TRUE}
bdffp <- read_csv(here("data_cleaned", "daily_precip.csv"), col_types = cols(site = col_character()))
skim(bdffp)
```

# Try as a tsibble

Going to try converting to a tsibble and using the tsibble package and other tidyverts packages for some analyses.

```{r}
bdffp_ts <- as_tsibble(bdffp, key = site, index = date)
```

## Visualize Gaps

This method only shows the *implicit* gaps (dates that are just not there in the spreadsheet)

```{r}
bdffp_ts %>% 
  count_gaps(.full = TRUE) %>% 
  ggplot(aes(y = site)) +
  geom_linerange(aes(xmin = .from, xmax = .to)) +
  geom_point(aes(x = .from)) +
  geom_point(aes(x = .to)) +
  labs(x = "Date", y = "Site", title = "Implicit gaps in data")
```

Here's one way to visualize **all** the gaps (includes dates entered into spreadsheet, but with no observations made)

```{r}
bdffp_ts %>% 
  filter(!is.na(precip)) %>% 
  count_gaps(.full = FALSE) %>% 
  ggplot(aes(y = site)) +
  geom_linerange(aes(xmin = .from, xmax = .to)) +
  # geom_point(aes(x = .from)) +
  # geom_point(aes(x = .to)) +
  labs(x = "Date", y = "Site", title = "All gaps in data")
```

But that's sort of useless.  


# Deal with accumulations

Mark possible untagged accumulations as any un-flagged precip value after one or more NAs.

```{r}
bdffp <- bdffp %>% 
  mutate(gaptype = case_when(
    flag == "accumulated" ~ "accumulated",
    !is.na(precip) & is.na(lag(precip)) ~ "untagged",
    !is.na(precip) & !is.na(lag(precip)) ~ "ordinary"
  ))

```

## Compare tagged, untagged, and no gap

How much rain after each type of gap?

```{r}
bdffp %>% 
  group_by(site) %>% 
  filter(!is.na(precip)) %>% 
  mutate(gaptype = fct_relevel(gaptype, "ordinary", "untagged", "accumulated")) %>% 
  ggplot(aes(precip)) +
  geom_histogram(aes(y = after_stat(ncount))) +
  facet_wrap("gaptype", ncol = 1) +
  labs(x = "Precipitation (mm)", y = "Proportion of observations", caption = '"ordinary" = no preceding gap\n "untagged" = preceding gap w/ no note\n "accumulated" = preceding gap w/ note that obs is accumulated precip')
```
I think at least some of the untagged values following gaps are probably accumulations, but the distributions are similar looking enough that I don't think it would be worth trying to establish a cutoff or something.

```{r}
bdffp %>% filter(gaptype == "untagged" & !is.na(notes))
bdffp %>% count(gaptype)
bdffp %>% filter(gaptype == "untagged", precip > 0) %>% count()
```
3360 potentially untagged accumulations

## Length of gaps

```{r}
# calculates number of NAs before an observation.  Excludes leading NAs at each site.
gaps <- 
  bdffp %>% 
  group_by(site) %>% 
  # mutate(row = row_number()) %>% #probably not necessary
  mutate(cumna = cumsum(is.na(precip))) %>% 
  filter(!is.na(precip)) %>% 
  mutate(gap_len = cumna - lag(cumna)) %>% 
  mutate(gap_len = na_if(gap_len, 0)) %>% 
  select(-cumna)
bdffp <- full_join(bdffp, gaps)  
```

```{r}
gap_summary <- 
  bdffp %>% 
  group_by(gaptype) %>% 
  summarize(mean_gap = mean(gap_len, na.rm = TRUE),
            median_gap = median(gap_len, na.rm = TRUE),
            max_gap = max(gap_len, na.rm = TRUE)) %>% 
  filter(gaptype %in% c("accumulated", "untagged"))
gap_summary
bdffp %>% filter(gap_len > 30 & flag == "accumulated")
```

Plotting gap length vs precip after gap.  (is this helpful?)

```{r}
bdffp %>% 
  filter(gaptype %in% c("accumulated", "untagged")) %>% 
  filter(gap_len < 300) %>% 
  ggplot(aes(x = gap_len, y = precip, color = gaptype)) +
  geom_point()
```

## Are NA's zeroes?

How are NAs distributed over time?  Are they more common in the dry season?  Then they might actually sometimes be zeroes (uh oh).

```{r}
bdffp %>% 
  mutate(month = month(date, label = TRUE)) %>% 
  group_by(site, month) %>%
  summarize(prop_NA = sum(is.na(precip))/n()) %>% 
  ggplot(aes(x = month, y = prop_NA)) + geom_col() +
  labs(x = "Month", y = "Proportion of missing values") +
  facet_wrap(~site) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# ggsave("distribution of NAs.png")
```

NAs are more common the the dry season in at least some sites indicating that some of those NAs are probably actually zeroes.


# Compare with Manaus Data

I'll remove any tagged or possible untagged accumulations, then compare remaining data with manaus...somehow.

```{r}
bdffp_ts <- 
  as_tsibble(bdffp, key = site, index = date) %>%
  fill_gaps()
bdffp_ts
```

```{r}
manaus <- read_csv(here("data_cleaned", "manaus_weather.csv"))
# manaus <- as_tsibble(manaus, index = date)
manaus <- manaus %>% filter(between(date, min(bdffp$date), max(bdffp$date)))
```

```{r}
ggplot(manaus, aes(x = date, y = precip)) + geom_line() + facet_wrap(~year(date), scales = "free_x")
```

```{r}
ggplot(bdffp, aes(x = date, y = precip, color = site)) + geom_line() + facet_wrap(~year(date), scales = "free_x")
```

```{r}
combined <-
  manaus %>%
  select(date, precip) %>% 
  add_column(site = "MANAUS") %>% 
  bind_rows(bdffp) %>% 
  # filter(is.na(flag)) %>% 
  filter(!gaptype %in% c("untagged", "accumulated")) %>% 
  as_tsibble(key = site, index = date) %>% 
  fill_gaps()
  
ggplot(combined %>% filter(year(date) == 2000), aes(x = date, y = precip, color = site)) +
  geom_line() +
  facet_wrap(~month(date), scales = "free_x")
```
```{r}
comb_wide <- combined %>% 
  select(date, site, precip) %>% 
  pivot_wider(date, names_from = site, values_from = precip) %>% 
  ungroup() %>% as.data.frame()
correlate(comb_wide %>% select(-date)) %>%
  shave() %>% 
  fashion()
```

Low correlations with Manaus station.
